The MACE is essentially unidimensional: Bifactor models of the `Maltreatment and Abuse Chronology of Exposure' scale

### 2022-07-30 manuscript/main.tex
manuscript updates (2022-07-30)

To assess the unidimensionality of the MACE scale, we also calculated the explained common variance (ECV; \citealt{sijtsma2009use}) and the proportion of uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale; it therefore measures the contribution of the general factor relative to the specific factors. As ECV values approach 1, the general factor loadings of a bifactor model are expected to resemble those that would be obtained by a one-dimensional model. ECV values greater than 0.70 often indicate a scale is essential unidimensional \citep{rodriguez2016applying}. In turn, the PUC quantifies how many inter-item correlations are accounted for only by the general factor and is an important moderator of the ECV. As the PUC increases, ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure. When PUC > 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated the relative parameter bias (RPB), which is the difference between an item's loading in a unidimensional model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.

@@

The MACE is an increasingly popular retrospective self-report measure of childhood maltreatment. Despite this, there has been little investigation into its factor structure \citep{saini2019systematic}, raising questions about the validity and interpretability of MACE subscale scores and composite indices. Here we employed a series of confirmatory and exploratory item factor models to investigate the dimensionality of the MACE. Simply put, we found repeated evidence across two independent samples that the MACE is essentially unidimensional. First, one-factor (unidimensional) models of the MACE provided adequate (albeit worse) fits to both datasets. Second, the omega hierarchical ($\omega_h$) for the full bifactor model was 0.924, indicating that the overwhelming majority of variance  in MACE severity scores is attributable to general maltreatment and only trivially affected by multidimensionality of the specific factors. Third, the explained common variance was, on average, approximately 0.70, suggesting that unidimensional models of the MACE are not expected to suffer from seriously biased parameter estimates.

As a corollary, we found little evidence for the reliability of MACE subscale scores or composite indices. Indeed, the omega subscale ($\omega_s$) values indicated that there was little reliable variance in almost all of the MACE subscale scores after accounting for the general factor. That is, MACE subscale scores cannot be interpreted as such but should instead be understood as predominantly reflecting overall maltreatment. The same can be said of the deprivation composite index, which had little reliable variance unique from the general factor and is seemingly contaminated by a reverse-scoring methods factor. Notably, we did find weak evidence for a secondary peer victimization factor in the MACE, whose composite scores possess a small majority of unique variance independent of general maltreatment. The existence of this peer victimization factor was supposed by exploratory factor modeling. Thus, it may be possible to derive two measures from the MACE, maltreatment by parents or other adults and maltreatment by peers, but further research is required to support this conclusion.

Our results are consistent with a number of studies investigating the factor structure of other childhood maltreatment measures. Bifactor models of the Childhood Trauma Questionnaire have similarly found a dominant general maltreatment factor with smaller specific factors \citep{spinhoven2014childhood, hollerbach2018main, stagaki2022mediating}. Similar results were obtained in recent applications of a bifactor model to the International Child Abuse Screening Tool \citep{meinck2021factor} and the Adverse Childhood Experience questionnaire \citep{dobson2021latent}. Together, these and our current results further highlight the high co-occurrence of childhood maltreatment experinces and the challenge that poses for investigating how particular forms of abuse shape development.

Our results are also worth considering alongside other recent studies using the MACE to investigate the link between particular types of childhood maltreatment and various psycho-biological outcomes. For example, some groups have looked at the association between MACE subscale scores and lifetime risk of depression \citep{gerke2018childhood}, dissociative symptoms \citep{schalinski2015type}, and cortisol concentration \citep{schalinski2019early}. Others have used the MACE to form composite indices of threat and deprivation experiences to study their respective links to cognitive functioning \citep{schalinski2018defining} and hippocampal volume \citep{teicher2018differential}. Our results are clear that the majority of reliable variance in these subscale and composite scores are attributable to overall maltreatment, thereby precluding meaningful interpretation of these measures of specific types of maltreatment as such. This in turn makes it difficult to make substantive sense of these studies highlighting the unique contribution of a subscale over the total severity score, given their redundancy, as well as the low reliability of the part of the subscale not redundant with the total score.

Worth further consideration is the peer victimization factor. Correlation between the bullying factor and the general factor was moderate. In accordance, a previous report on two prospective cohorts estimated that around 40\% of children who were maltreated were also bullied \citep{lereya2015adult}. The authors of that study speculated that there may be common risk factors for childhood maltreatment and bullying such as family instability and poor social skills. On the other hand, the authors also hypothesized that maltreatment may make children more susceptible to bullying by, for instance, disrupting the child's ability to regulate emotions. Similar to childhood maltreatment by household members, bullying by peers is a common childhood experience \citep{world2020spotlight} and has detrimental effects on mental health independent of other types of childhood maltreatment \citep{singham2017concurrent, lereya2015adult}. Separating peer bullying as its own factor in the MACE may therefore offer a more granular view of the harmful childhood experiences of the responders. In the future, the peer bullying factor should be validated against other restrospective bullying measures such as the California Bullying Victimization Scale \citep{green2018initial}. Finally, the MACE does not assess cyberbullying which is an increasingly common form of bullying.

As part of current analyses, we also discovered other features of the MACE. We identified sources of differential item functioning between men and women. Across datasets, women endorsed more sexual abuse, and men reported more physical abuse by peers. These results are in line with previous studies showing that girls are more often victims of sexual abuse \citep{stoltenborgh2015prevalence}, whereas boys are more often victims of bullying, especially physical bullying \citep{scheithauer2006physical}. We also identified a number of other items that exhibited some differential item functioning across samples. These differences may be attributable to different populations (Boston vs. across Canada \& USA), recruitment (local advertisement vs. prolific.co), exclusion criteria (required to be healthy and unmedicated vs. none), generational effect (data collected in 2010-2013 vs. 2021), technical differences in the online assessment platform ("HIPAA-compliant online enrollment system" vs. website built specifically to collect MACE data with jspsych \citep{de2015jspsych}), or differences in excluding inattentive or careless responders (none vs. infrequency items \citep{zorowitz2021inattentive}). The causes and consequences of measurement invariance in the MACE deserves further study.

We also identified others methods features of the MACE. For one thing, we observed a consistent reverse-scoring methods factor in the MACE. One possibility is that this is driven by response styles, as has been seen in other questionnaires. A second possibility is that this is induced by the scoring procedure. Future studies should investigate alternative scoring or remove them altogether. A second thing, the MACE has a large number of items with response dependence. That is, the MACE has far fewer than 52 independent items. It is possible then that the MACE could be abbreviated, though this may lose sensitivity by dropping items with high/low difficulty. Again, future work.

The present study is, of course, not without its own limitations. One obvious limitation is that we only study the binary endorsement data, but do not investigate the chronology data also sampled as part of the MACE. Though the endorsement data is predominantly unidimensional, this may belie other important patterns present only in the chronology data. In actuality, there are two dimensions governing responses in the MACE: susceptibility to maltreatment, or how many types or forms of maltreatment a person experiences, and severity, or how much or how long a person experienced maltreatment. Such a process may be well modeled by a multidimensional hurdle response model \citep{magnus2021symptom}, modeling a person's probability of endorsing a maltreatment event and, if endorsed, the frequency of the maltreatment. This would allow study of the independence of these two dimensions, but could also provide insight into the structural validity of the chronology data. Fodder for future work.

A second limitation is the sample. Though our sample studied responses in the MACE in a more diverse sample with respect to gender and age, our sample is still lacking. Our sample is not racially or ethnically diverse. To the extent to which other studies have identified measurement invariance in more diverse samples (https://doi.org/10.1016/j.jpsychores.2007.04.010, https://doi.org/10.1016/j.chiabu.2019.104030), it is not clear to what extent the factor structure explored here holds. More research is necessary to ensure that the MACE is valid cross-culturally.

To conclude, we found that the MACE is essentially unidimensional. Maltreatment endorsements are, for the most part, governed by a unitary maltreatment susceptibility factor. The MACE total score is valid and reliable. The MACE subscale scores, however, are not and should not be treated as interpretable measures of maltreatment subtypes. Researchers should be careful not to use subscale scores before they are validated, lest they draw spurious inferences.

>>>>

To assess the unidimensionality of the MACE scale, we also calculated the explained common variance (ECV; \citealt{sijtsma2009use}) and the proportion of uncontaminated correlations (PUC; \citealt{reise2013multidimensionality}). The ECV is the ratio of the variance explained by the general factor divided by the total common variance of a scale; it therefore measures the contribution of the general factor relative to the specific factors. As ECV values approach 1, the general factor loadings of a bifactor model are expected to resemble those that would be obtained by a one-dimensional model. ECV values greater than 0.70 often indicate a scale is essentially unidimensional \citep{rodriguez2016applying}. In turn, the PUC quantifies how many inter-item correlations are accounted for only by the general factor and is an important moderator of the ECV. As the PUC increases, ECV is less important for evaluating the risk of bias when fitting a unidimensional model to data with a latent bifactor structure. When PUC > 0.80, there is a low risk of bias when a multidimensional scale is treated as unidimensional \citep{reise2013multidimensionality}. As an additional check, we calculated the relative parameter bias (RPB), which is the difference between an item's loading in a unidimensional model and its corresponding general factor loading the bifactor, divided by the general factor loading in the bifactor. Parameter bias less than 10–15\% is acceptable \citep{muthen1987structural}.

@@

The MACE is a promising alternative measure of childhood maltreatment \citep{georgieva2022systematic}. However, there has been little study of its structural validity \citep{saini2019systematic}, leaving unanswered questions about the scale's underlying factor structure and the validity of its total/subscale scores. Here we investigated the factor structure of the MACE using both confirmatory and exploratory item response modeling. The fits of a bifactor model (model 2) to the response data from both samples revealed a dominant general factor with uniformly strong loadings from the majority of items, and substantially weaker loadings on the specific subscale factors. The general maltreatment factor explained approximately 70\% of the reliable variance in responses across samples. When comparing the loadings on the general factor to those from a unidimensonal model, the relative parameter bias was marginal. Together, these results support the conclusion that the MACE is essentially unidimensional.

We also calculated model-based reliability indices of the MACE total and subscale/composite scores using the bifactor model fits. We found that the MACE total score has excellent reliability and is a valid measure of general maltreatment. In contrast, we found that the MACE subscale scores were not valid measures of specific forms of maltreatment; in all but one example, the majority of reliable variance in subscale scores was attributable to the general factor, thereby precluding meaningful interpretation of the MACE subscale scores as such. The same was true of the composite deprivation scores, derived from the neglect items on the MACE, which also possessed little reliable variance independent of the general factor and was contaminated by a reverse-scoring methods factor. In sum, whereas the MACE total score is reliable and valid, composite scores derived from a subset of items are seldom interpretable as such and instead reflect general maltreatment.

Of the four confirmatory factor models we tested, the best-fitting and most parsimonious was a bifactor S-1 model with dominant general maltreatment factor and two smaller specific factors reflecting peer victimization and reverse-scoring (model 4). This model of responses on the MACE was further supported by the exploratory factor analyses, where the 3-factor solution found a similar partitioning of the items. Notably, across samples, composite scores derived from the peer victimization items were both reliable and interpretable insofar that the majority of their variance were attributable to the peer victimization factor even after accounting for the general maltreatment factor. It therefore may be valid to derive two score measures from the MACE, maltreatment by adults and by peers, but further research is needed to properly validate these measures.

Our results are consistent with a number of previous studies investigating the factor structure of other childhood maltreatment measures. For example, bifactor models of the Childhood Trauma Questionnaire have also revealed a dominant general maltreatment factor and weaker specific factors \citep{spinhoven2014childhood, hollerbach2018main, stagaki2022mediating}. Similar results were found in recent applications of a bifactor model to the International Child Abuse Screening Tool \citep{meinck2021factor} and the Adverse Childhood Experience questionnaire \citep{dobson2021latent}. Together, these and our current results underscore the challenge of developing versatile measures of childhood maltreatment. When the experience of distinct forms of abuse are correlated, it is difficult to measure any particular type of abuse independent of overall maltreatment.

Our results are also worth considering alongside previous studies that used subscale and/or composite scores from the MACE to investigate the link between specific forms of childhood maltreatment and assorted psycho-biological outcomes. For example, researchers have looked at the relationship between MACE subscale scores and risk of depression \citep{gerke2018childhood}, dissociative symptoms, \citep{schalinski2015type}, and cortisol concentration \citep{schalinski2019early}. Still others have derived composite threat and deprivation scores from the MACE to study their associations with cognitive functioning \citep{schalinski2018defining} and hippocampal volume \citep{teicher2018differential}. Our results are unequivocal that these measures principally reflect general maltreatment and not particular forms of adversity. This presents a challenge for interpreting the results of those studies. It is worth remarking that several of these studies attempted to overcome these issues by simultaneously included all of the MACE subscale/composite scores in their models using more sophisticated multivariate analyses (e.g. random forest regression). This approach is unsatisfactory for two reasons. First, random forest models can return spurious results when handling collinear predictors \citep{gregorutti2017correlation}. Second, even if one were to properly control for general maltreatment, our results indicate that the remaining reliable variance in subscale scores is so low as to render any inference of questionable utility.

%Worth further consideration is the peer victimization factor. Correlation between the bullying factor and the general factor was moderate. In accordance, a previous report on two prospective cohorts estimated that around 40\% of children who were maltreated were also bullied \citep{lereya2015adult}. The authors of that study speculated that there may be common risk factors for childhood maltreatment and bullying such as family instability and poor social skills. On the other hand, the authors also hypothesized that maltreatment may make children more susceptible to bullying by, for instance, disrupting the child's ability to regulate emotions. Similar to childhood maltreatment by household members, bullying by peers is a common childhood experience \citep{world2020spotlight} and has detrimental effects on mental health independent of other types of childhood maltreatment \citep{singham2017concurrent, lereya2015adult}. Separating peer bullying as its own factor in the MACE may therefore offer a more granular view of the harmful childhood experiences of the responders. In the future, the peer bullying factor should be validated against other restrospective bullying measures such as the California Bullying Victimization Scale \citep{green2018initial}. Finally, the MACE does not assess cyberbullying which is an increasingly common form of bullying.

The current study revealed other features of the MACE worth remarking on. For example, we identified several sources of DIF on the MACE. Across samples we found that women were more likely to endorse having been sexually abused, whereas men were more likely to report being physically abused by peers. These findings are in line with previous studies that found girls are more frequently victims of sexual abuse \citep{stoltenborgh2015prevalence} and boys are more frequently victims of bullying (especially physical bullying; \citealt{scheithauer2006physical}). We also identified a number of items that exhibited DIF across samples. The cause of this is unclear, and may reflect differences in sample population (greater Boston area vs. US \& Canada), study location (in clinic vs. online), recruitment years (early 2010s vs. early 2020s), inclusion criteria (healthy \& unmedicated vs. none), and/or exclusion criteria (none vs. attention checks). The causes and impact of violations of measurement invariance on the MACE deserve further study.

We also identified a reverse-scoring methods factor in the MACE. In our view, the most plausible explanation is that, under the \cite{teicher2015maltreatment} scoring procedure, affirmative responses on the regular and reverse-scored items on the MACE represent very different responses. For a regular item on the MACE, an endorsement indicates that a person experienced an adverse event during one year of childhood at least. In contrast, an endorsement on a reverse-scored item indicates that a person experienced experienced the absence of a positive event for all 18 years of childhood. Alternative causes of the reverse-scoring factor, such as participant confusion, acquiescence, or carelessness \citep{weijters2013reversed}, are also possible. Additional research is needed to see if alternative scoring procedures eliminates this factor.

The current study is, of course, not without limitations. The most apparent limitation is that here we only studied the dichotomous response data but not the chronology data also solicited as part of the MACE. Though we found that the factor structure of the dichotomous responses are essentially unidimensional, this may belie more complex patterns present in the maltreatment onset and duration data. Indeed, modeling the timeseries of adverse events in childhood --- perhaps using dynamic factor analysis \citep{zhang2007bayesian} --- might be more sensitive to more nuanced patterns of maltreatment co-occurrence that are lost when collapsing the data into a measure of any vs. no maltreatment. Another possibility is to jointly model the MACE endorsement and chronology data using a multidimensional hurdle model \citep{magnus2021symptom}, which measures both a person's susceptibility to maltreatment (i.e. number of unique types of maltreatment experienced) and severity of maltreatment (i.e. for how long maltreatment was experienced). More research is needed to develop valid and reliable MACE scores that integrate both the endorsement and chronology data.

A second limitation concerns our sample. Though the inclusion of the replication sample increased the overall diversity of our participants with respect to gender and age, other aspects of diversity are notably lacking. For example, our combined sample was not racially or ethnically diverse. This is important as previous studies have identified DIF for childhood trauma questionnaires by race \citep{thombs2007evaluation, rodriguez2019identification}, which raises the possibility that similar issues could be found for the MACE with more diverse samples. Moreover, as noted above, we are not currently able to explain why DIF was identified between the original and replication samples. Further research is needed to identify and unpack violations of measurement invariance in the MACE in order to effectively assess childhood maltreatment in many different populations.

To conclude, we found that the factor structure of the MACE is essentially unidimensional. Despite the presence of smaller secondary factors, bifactor modeling revealed that the large majority of reliable variance in the dichotomous MACE responses are attributable to a general maltreatment factor. Thus, we found that the MACE total score is a valid and reliable of overall childhood maltreatment. In contrast, we found that MACE subscale and composite scores are not valid measures of particular types of childhood maltreatment, but instead primarily reflect general maltreatment. Moving forward, childhood abuse and neglect researchers utilizing the MACE (or otherwise) should take pains to ensure that any subscale or composite scores they employ are valid, lest they draw spurious inferences.
